---
title: 'Section 6: Data Classes'
author: 'MTH 365: Introduction to Data Science'
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

## Data types in R

R is (usually) good at figuring out the best data type, but sometimes we'll need to take corrective action! The five data types we'll interact with the most are:

- Logical
- Double
- Integer
- Character
- Lists

```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
```

__Logical__: 

```{r}
typeof(TRUE)
```

__Character__: 

```{r}
typeof("hello")
typeof('world')
```

__Double__: 

```{r}
typeof(1.335)
typeof(7)
```

__Integer__: 

```{r}
typeof(7L)
typeof(1:3)
```

__Lists__:

```{r}
mylist <- list("A", 1:4, c(TRUE, FALSE), (1:4)/2)
mylist

str(mylist)
```

The `:` symbol generates "+1" lists: these have many uses, including iteration and specifying starting values.

```{r}
1:5
typeof(1:5)
1.1:5.1
typeof(1.1:5.1)
```

Another useful type of lists are "sequences". These can increment by any amount.

```{r}
seq(from=1, to=5, by=1)
seq(from=1, to=5)
seq(from=1, to=5, by=2)
```

## Named lists

Because of their more complex structure we often want to name the elements of a list (we can also do this with vectors). This can make reading and accessing the list more straight forward.

```{r}
myotherlist <- list(A = "hello",
                    B = 1:4, 
                    "knock knock" = "who's there?")
str(myotherlist)

names(myotherlist)
myotherlist$B
```

## Concatenation

Vectors can be constructed using the `c()` function.

```{r}
c(1, 2, 3)
c("Hello", "World!")
c(1, c(2, c(3)))
```

- Default type for numeric vectors is:

## Coercion

When you mix data types within a vector, R will create the result with the data type that can most easily accommodate all the elements it contains. This conversion between modes of storage is called “coercion”. 

- When R converts the mode of storage based on its content, it is referred to as “implicit coercion”. Can you guess the following data types (without running them first)?

```{r}
c(1, "Hello")
c(FALSE, 3L)
c(1.2, 3L)
```

Use the `typeof()` function to find out.

## Missing values

R uses `NA` to represent missing values in its data structures.

```{r}
typeof(NA)
```

## Other special values

- `NaN`: Not a number
- `Inf`: Positive infinity
- `-Inf`: Negative infinity

```{r}
pi
pi / 0
0 / 0

1/0 + 1/0
1/0 - 1/0
NaN / NA
NaN * NA
```

## Practice

\textcolor{red}{Example}: What is the type of the following vectors? Explain why they have that type.

- `c(1, NA+1L, "C")`
- `c(1L / 0, NA)`
- `c(1:3, 5)`
- `c(3L, NaN+1L)`
- `c(NA, TRUE)`


## Example: Cat lovers

A survey asked respondents their name and number of cats. The instructions said to enter the number of cats as a numerical value.

```{r message=FALSE, echo=FALSE}
# Make sure to change the file directory
cat_lovers <- read.csv("~/OneDrive - Creighton University/teaching-courses/creighton/Data Science course content/original/Extra Files/cat_lovers.csv")
```

```{r}
glimpse(cat_lovers)
```

Oh why won't you work?!

```{r, error=TRUE}
cat_lovers %>%
  summarise(mean = mean(number_of_cats))
```

Seriously?

```{r}
cat_lovers %>%
  summarise(mean_cats = mean(number_of_cats, na.rm = TRUE))
```

Slow down.

What is the type of the `number_of_cats` variable?
  
```{r}
typeof(cat_lovers$number_of_cats)
```

- Who answered "three"?

```{r echo=FALSE}
summary(cat_lovers)
```

Are there any other strange responses?

```{r} 
cat_lovers %>% filter(number_of_cats=='three')

cat_lovers %>% filter(number_of_cats=='1.5 - honestly I think one of my cats is half human')
```

This is why surveys restrict entry values...

```{r}
cat_lovers %>%
  mutate(
    number_of_cats = case_when(
      name == "Ginger Clark" ~ "2",
      name == "Doug Bass"    ~ "3"
    ),
    number_of_cats = as.numeric(number_of_cats)
  ) %>%
  summarise(mean_cats = mean(number_of_cats))
```

Try something: define a new variable?

```{r}
cat_lovers2 <- cat_lovers %>%
  mutate(
    number_of_cats = case_when(
      name == "Ginger Clark" ~ "2",
      name == "Doug Bass"    ~ "3"
    ),
    number_of_cats2 = as.numeric(number_of_cats)
  ) %>%
  summarise(mean_cats = mean(number_of_cats))
```

Crap. Spot the problem.

```{r}
glimpse(cat_lovers2)
```

Now that we (maybe) know what we're doing...

```{r}
cat_lovers %>%
  mutate(number_of_cats = case_when(
    name == "Ginger Clark" ~ 2,
    name == "Doug Bass"    ~ 3,
    TRUE                   ~ as.numeric(number_of_cats)
  )) %>%
  summarise(mean_cats = mean(number_of_cats))
```



## Moral of the story

- If your data doesn't behave how you expect it to, implicit coercion a might be the reason.
- Go in and investigate your data, apply the fix, _save your data_, live happily ever after.

## Vectors vs. lists

```{r}
x <- c(8,4,7)
```

```{r}
x[1]
```

```{r}
x[[1]]
```

```{r}
y <- list(8,4,7)
```

```{r}
y[2]
```

```{r}
y[[2]]
```

**Note:** When using tidyverse code you'll rarely need to refer to elements using square brackets, but it's good to be aware of this syntax, especially since you might encounter it when searching for help online.


## Data "sets" in R

- "set" is in quotation marks because it is not a formal data class

A tidy data "set" can be one of the following types:

- `tibble`
- `data.frame`

We'll often work with `tibble`s:

- `readr` package (e.g. `read_csv` function) loads data as a `tibble` by default
- `tibble`s are part of the tidyverse, so they work well with other packages we are using
- They make minimal assumptions about your data, so are less likely to cause hard to track bugs in your code


## Data frames

A data frame is the most commonly used data structure in R, they are just a list of equal length vectors (usually atomic, but you can use generic as well). 

- Each vector is treated as a column and elements of the vectors as rows.

A `tibble` is a type of data frame that ... makes your life (i.e. data analysis) easier.

- Most often a data frame will be constructed by reading in from a file, but we can also create them from scratch.

```{r}
df <- tibble(x = 1:3, y = c("a", "b", "c"))
class(df)
glimpse(df)

attributes(df)
class(df$x)
class(df$y)
```

Example: How many respondents have below average number of cats?

```{r}
cat_lovers<-cat_lovers %>%
  mutate(number_of_cats = case_when(
    name == "Ginger Clark" ~ 2,
    name == "Doug Bass"    ~ 3,
    TRUE                   ~ as.numeric(number_of_cats)
  ))
```


```{r}
mean_cats <- cat_lovers %>%
    summarise(mean_cats = mean(number_of_cats))
cat_lovers %>%
    filter(number_of_cats < mean_cats) %>%
    nrow()
```

Do you believe this number? Why or why not? (Hint: How many respondents were there?)

```{r}
mean_cats
class(mean_cats)
```

 A possible solution: `pull()`.

`pull()` works like [[]] for data frames, and pulls out the value of a single variable in a data frame.

```{r}
data(mtcars)
glimpse(mtcars)
```

How does `pull()` work?

```{r}
mtcars %>% pull()
mtcars %>% pull(2)
mtcars %>% pull(cyl)
```

`pull()` can be your new best friend, when used correctly.

```{r,}
mean_cats <- cat_lovers %>%
summarise(mean_cats = mean(number_of_cats)) %>%
pull()
cat_lovers %>%
filter(number_of_cats < mean_cats) %>%
nrow()
```

```{r, eval=FALSE}
mean_cats
class(mean_cats)
```

## Factors

Factor objects are how R stores data for categorical variables (fixed numbers of discrete values).

```{r}
(x = factor(c("BS", "MS", "PhD", "MS")))
```

```{r}
glimpse(x)
```

```{r}
typeof(x)
```

What if we want to force a particular order?

```{r}
glimpse(cat_lovers)

ggplot(cat_lovers,aes(x = handedness)) +
  geom_bar()
```

## `forcats`

```{r}
cat_lovers <- cat_lovers %>%
mutate(handedness = fct_relevel(handedness, 
"left", "ambidextrous", "right"))

ggplot(cat_lovers, aes(x = handedness)) +
  geom_bar(aes(fill=handedness))
```

## Factor notes

- R uses factors to handle categorical variables, variables that have a fixed and known set of possible values. Historically, factors were much easier to work with than character vectors, so many base R functions automatically convert character vectors to factors.

- However, factors are still useful when you have true categorical data, and when you want to override the ordering of character vectors to improve display. The goal of the forcats package is to provide a suite of useful tools that solve common problems with factors.

Source: [forcats.tidyverse.org](http://forcats.tidyverse.org/)

## Recap

Always best to think of data as part of a tibble

  - This plays nicely with the `tidyverse` as well
  - Rows are observations, columns are variables

Be careful about data types/classes

  - Sometimes `R` makes silly assumptions about your data class 
  - Using `tibble`s help, but it might not solve all issues
  - Think about your data in context, e.g. 0/1 variable is most likely a `factor`
  - If a plot/output is not behaving the way you expect, first investigate the data class
  - If you are absolutely sure of a data class, overwrite it in your tibble so that you don't need to keep having to keep track of it
  - `mutate` the variable with the correct class
  
  
  Reflection Questions:

1.  How did the group work in Zoom go last week?  Do you think its worth continuing to do? 

2.  Are there any topics you specifically want to address during the class meeting?

3.  Any additonal thoughts/comments on the notes:
  