---
title: "Section 2: Data Visualization (Part 1)"
author: 'MTH 365: Introduction to Data Science'
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

## Recommended Reading

- _Modern Data Science with R_ Ch. 2: Data Visualization
- _Modern Data Science with R_ Ch. 3: A Grammar for Graphics

```{r, message=FALSE}
#install.packages(dslabs)
library(dslabs)
```

## `ggplot2`

`ggplot2` is an R package for "declaratively creating graphics"

- https://ggplot2.tidyverse.org/reference/

__Basic Format of a Plot__:

`data and aesthetics + plot type + options`

```{r, message=FALSE}
#install.packages(tidyverse)
library(tidyverse)
library(ggplot2)
```

## Example: Hate crimes and income inequality

A FiveThirtyEight article published in 2017 claimed that higher rates of hate crimes were tied to greater income inequality.

https://fivethirtyeight.com/features/higher-rates-of-hate-crimes-are-tied-to-income-inequality/

- FiveThirtyEight publishes their data sets - let's investigate for ourselves.
- Data sets are posted on BlueLine. Download this data set, and save it to your computer. 

Follow these steps to read the data into RStudio:

1. In the Environment tab, click "Import Dataset". Since this is a CSV document, you want to import a text file.
2. Navigate to your CSV data set. Make sure that the first row contains column names. 
3. Import the data.

You'll see a line of code pop up in your R console, it should start with `read.csv()`. Copy this code and replace the provided statement (`hate_crimes <- read.csv("~/OneDrive - Creighton University/Spring 2019 Courses/MTH 365 - Intro to Data Science/Data/hate_crimes.csv")`) with the one you just generated.

```{r}
library(readr)
hate_crimes <- read_csv("~/School/MTH 365/hate_crimes.csv")
glimpse(hate_crimes)
```

## A simple plot?

```{r}
ggplot(hate_crimes, aes(x=median_income, y=hate_crimes_fbi)) +
  geom_point()
```

```{r}
ggplot(hate_crimes, aes(x=median_income, y=hate_crimes_fbi))
```

```{r}
ggplot(hate_crimes, aes(x=median_income, y=hate_crimes_fbi)) + 
  geom_text(aes(label=state))
```

```{r}
ggplot(hate_crimes, aes(x=median_income, y=hate_crimes_fbi)) +
  geom_text(aes(label=state, color=region))
```

## Measuring income inequality

```{r}
ggplot(hate_crimes, aes(x=gini_index, y=hate_crimes_fbi)) +
  geom_point(aes(color=region))
```

Does source matter?

```{r}
ggplot(hate_crimes, aes(x=gini_index, y=hate_crimes_splc)) +
  geom_point(aes(color=region))
```

## Adding plot options

```{r}
ggplot(hate_crimes, aes(x=gini_index, y=hate_crimes_fbi)) +
  geom_point(aes(color=region)) + 
  xlab('Gini Index') + 
  ylab('Hate Crimes (FBI, Sept. 2016)')
```

```{r}
ggplot(hate_crimes, aes(x=gini_index, y=hate_crimes_fbi)) +
  geom_point(aes(color=region, size=unemployment)) + 
  xlab('Gini Index') + 
  ylab('Hate Crimes (FBI, Sept. 2016)')
```

```{r}
ggplot(hate_crimes, aes(x=gini_index, y=hate_crimes_fbi)) +
  geom_point(aes(color=region, size=unemployment)) + 
  xlab('Gini Index') + 
  ylab('Hate Crimes (FBI, Sept. 2016)') + 
  facet_wrap(~region, nrow=1)
```

## Weather patterns

The data set `Weather` contains data on weather-related variables for several world cities.

```{r}
#install.packages(mosaicData)
library(mosaicData)
data(Weather)
glimpse(Weather)
```

```{r}
ggplot(Weather, aes(y=avg_temp)) + 
  geom_boxplot()
```

```{r}
ggplot(Weather, aes(y=avg_temp, group=city)) + 
  geom_boxplot(aes(color=city))
```

```{r}
ggplot(Weather, aes(x=date, y=avg_temp)) +
  geom_point()
```

```{r}
ggplot(Weather, aes(x=date, y=avg_temp)) +
  geom_point(aes(color=city))
```

### Weather patterns in Chicago

```{r}
Chicago <- Weather %>% filter(city=='Chicago')
ggplot(Chicago, aes(x=date, y=avg_temp)) +
  geom_point(col='blue')
```

```{r}
ggplot(Chicago, aes(x=date, y=avg_temp)) +
  geom_point(aes(color=events))
```

```{r}
ggplot(Chicago, aes(x=date, y=avg_temp)) +
  geom_smooth()
```

```{r}
ggplot(Chicago, aes(x=date, y=avg_temp)) +
  geom_smooth() +
  geom_point()
```

```{r}
ggplot(Chicago, aes(x=low_temp, y=high_temp)) +
  geom_point(aes(color=month))
```

```{r}
ggplot(Chicago, aes(x=low_temp, y=high_temp)) +
  geom_point(aes(color=as.factor(month)))
```

```{r}
ggplot(Chicago, aes(x=avg_temp)) + 
  geom_histogram()
```

```{r}
ggplot(Chicago, aes(x=avg_temp)) + 
  geom_density(fill='lightblue')
```

```{r}
ggplot(Chicago, aes(x=avg_temp)) + 
  geom_density(fill='lightblue', alpha=0.5)
```

```{r}
ggplot(Chicago, aes(x=avg_temp, group=month)) + 
  geom_density(aes(color=as.factor(month), fill=as.factor(month)), alpha=0.5)
```

```{r}
ggplot(Chicago, aes(x=avg_temp, group=month)) +
  geom_density(aes(color=as.factor(month), fill=as.factor(month)), alpha=0.5) + 
  facet_wrap(~month, nrow=3)
```

```{r}
ggplot(Chicago, aes(x=avg_vis, group=month)) + 
  geom_density(aes(color=as.factor(month), fill=as.factor(month)), alpha=0.5) +
  facet_wrap(~month, nrow=3)
```

## Weather events

```{r}
ggplot(Weather, aes(x=events)) +
  geom_bar()
```

```{r}
ggplot(Weather, aes(x=events)) +
  geom_bar() +
  coord_flip()
```

### Weather events in Chicago

```{r}
ggplot(Chicago, aes(x=events)) +
  geom_bar() +
  coord_flip() +
  facet_wrap(~year, nrow=2)
```

## Mutate Command

We're going to cover one data wrangling command here, and its the 'Mutate' command.  The mutate function is used to add columns to a dataframe (and not, as the same seems to suggest, change existing columns).  As an example, we are going to add a column to the Weather dataset that records the gap between the high temperature and the low temperature for each row.

```{r}
weather_gap<-Weather%>%
  mutate(gap =high_temp-low_temp)
glimpse(weather_gap)
```

Here, we're naming our new dataframe weather_gap.  The mutate function requires a dataframe as its first entry (which here is the Weather dataset that we are piping in).  The "gap=high_temp-low_temp" line means to name the new column "gap", and that new column is found by subtracting the high_temp from the low_temp.  After running the "glimpse" command, you can see the new column in the dataframe.

In this code chunk, try mutatins a categorial data column using algebraic operations.  Below this chunk, explain what happened, as best you understand it.
```{r}

```

Reflection Questions:

1.  With 1 being too slow and 5 being too fast, how was the speed of the video? 
  3, the video was a good pace

2.  With 1 being too basic and 5 being too advanced, how was the level of the video? 
  4 it was hard to comprehend initially but I got it after I talked to the prof and watched it again.

3.  Are there any topics you specifically want to address during the class meeting?
  Nope
  
4.  Any additonal thoughts/comments on the notes:
  Now that I'm not overthinking them, it's definately easier