---
title: "Lab 3: Choosing a Major"
subtitle: "MTH 365: Intro to Data Science"
author: "Your Name Here"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

## Before we begin, a "programming note":

<span style="color:red">
If you haven't yet, I _strongly consider_ giving font color changes a try for your written responses.  
</span>

<span style="background-color:yellow">
You can change the text color _or_ the background color! Or both! 
</span>

<span style="background-color:#c3dbf7">
Either way, formatting and style changes can really make your work (1) more readable, and (2) more "professional"-looking.
</span>

More info: https://www.bitdegree.org/learn/html-color-tag

- There are other ways to use HTML to change the font color -- whatever floats your boat!

-----

In this lab you'll explore data on college majors and earnings, specifically the data begind the FiveThirtyEight story ["The Economic Guide To Picking A College Major"](https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/).

These data originally came from the American Community Survey (ACS) 2010-2012 Public Use Microdata Series. While this is outside the scope of this lab, if you are curious about how raw data from the ACS were cleaned and prepared, see [the code](https://github.com/fivethirtyeight/data/blob/master/college-majors/college-majors-rscript.R) FiveThirtyEight authors used.

We should also note that there are many considerations that go into picking a major. Earnings potential and employment prospects are two of them, and they are important, but they don't tell the whole story. Keep this in mind as you analyze the data.

Start by loading the usual packages with the `library(tidyverse)` command. The data is in the `fivethirtyeight` package - you might need to install this one.

```{r}
library(tidyverse)
# install.packages(`fivethirtyeight`)
library(fivethirtyeight)
```

# Load the data

The data frame we will be working with today is called `college_recent_grads` and it's in the `fivethirtyeight` package.

To find out more about the dataset, type the following in your Console: `?college_recent_grads`. A question mark before the name of an object will always bring up its help file. This command must be ran in the Console.

`college_recent_grads` is a tidy **data frame**, with each row 
representing an **observation** and each column representing a **variable**.

To view the data, click on the name of the data frame in the Environment tab.

You can also take a quick peek at your data frame and view its dimensions with the `glimpse` function.

```{r glimpse}
glimpse(college_recent_grads)
```

The description of the variables, i.e. the **codebook**, is given below.

Header|Description
---|---
`rank`|Rank by median earnings
`major_code`|Major code, FO1DP in ACS PUMS
`major`|Major description
`major_category`|Category of major from Carnevale et al
`total`|Total number of people with major
`sample_size`|Sample size (unweighted) of full-time, year-round ONLY (used for earnings)
`men`|Male graduates
`women`|Female graduates
`sharewomen`|Women as share of total
`employed`|Number employed (ESR == 1 or 2)
`employed_full_time`|Employed 35 hours or more
`employed_part_time`|Employed less than 35 hours
`employed_full_time_yearround`|Employed at least 50 weeks (WKW == 1) and at least 35 hours (WKHP >= 35)
`unemployed`|Number unemployed (ESR == 3)
`unemployment_rate`|Unemployed / (Unemployed + Employed)
`median`|Median earnings of full-time, year-round workers
`p25th`|25th percentile of earnigns
`p75th`|75th percentile of earnings
`college_jobs`|Number with job requiring a college degree
`non_college_jobs`|Number with job not requiring a college degree
`low_wage_jobs`|Number in low-wage service jobs

The `college_recent_grads` data frame is a trove of information. Let's think about some questions we might want to answer with these data:

- Which major has the lowest unemployment rate?
- Which major has the highest percentage of women?
- How do the distributions of median income compare across major categories?
- Do women tend to choose majors with lower or higher earnings?

In the next section we aim to answer these questions.

# Data wrangling and visualization

## Which major has the lowest unemployment rate?

> 1. Create a new data set, `major_unemployment` with the following variables: `unemployment_rate`, `major`, `major_code`.

In this data set, the unemployment rate is reported to nine decimal places. This is probably a bit much! We can use `mutate()` and `round()` to change the number of decimal places of a variable.

```{r}
# college_recent_grads <- college_recent_grads %>%
#  mutate(unemployment_rate = round(unemployment_rate, digits = 4))
```

> 2. `mutate()` the `unmployment_rate` variable to three decimal places.

> 3. Which major has the _lowest_ unemployment rate?

> 4. Which major has the _highest_ unemployment rate?


## Which major has the highest percentage of women?

> 5. Arrange the data in descending order with respect to proportion of women in a major, and display only the major, the total number of people with major, and proportion of women. Show only the top 4 majors by adding `head(4)` at the end of the pipeline.

## How do the distributions of median income compare across major categories?

A percentile is a measure used in statistics indicating the value below which a given percentage of observations in a group of observations fall. For example, the 20th percentile is the value below which 20% of the observations may be found. (Source: [Wikipedia](https://en.wikipedia.org/wiki/Percentile)

There are three types of incomes reported in this data frame: `p25th`, `median`, and `p75th`. These correspond to the 25th, 50th, and 75th percentiles of the income distribution of sampled individuals for a given major.

> 6. Why do we often choose the median, rather than the mean, to describe the typical income of a group of people?

The question we want to answer is "How do the distributions of median income compare across major categories?". We need to do a few things to answer this question: First, we need to group the data by `major_category`. Then, we need a way to summarize the distributions of median income within these groups. This decision will depend on the shapes of these distributions. So first, we need to visualize the data.

> 7. Make a histogram of all median incomes, without splitting into categories.

Along with the plot, you might get a message:

`stat_bin()` using `bins = 30`. Pick better value with `binwidth`.

- This is telling us that we might want to reconsider the binwidth we chose for our histogram -- or more accurately, the binwidth we didn't specify. It's good practice to always thing in the context of the data and try out a few binwidths before settling on a binwidth. You might ask yourself: "What would be a meaningful difference in median incomes?" \$1 is obviously too little, \$10000 might be too high.

> 8. Try binwidths of \$1000 and \$5000. Choose one, and explain your reasoning for your choice. 

We can also calculate summary statistics for this distribution using the `summarise` function:

```{r}
college_recent_grads %>%
  summarise(min = min(median), max = max(median),
            mean = mean(median), med = median(median),
            sd = sd(median), 
            q1 = quantile(median, probs = 0.25),
            q3 = quantile(median, probs = 0.75))
```

- Some references use `summarize` and some use `summarise`. These are identical! US spellings and Aussie spellings both work in the `tidyverse`.

> 9. Based on the shape of the histogram you created in the previous exercise, determine which of these summary statistics is useful for describing the distribution. Write up your description (remember shape, center, spread, any unusual observations) and include the summary statistic output as well.

> 10. Plot the distribution of `median` income using a histogram, faceted by `major_category`. Use the `binwidth` you chose in the earlier exercise.

> 11. Plot the distribution of `median` income using a density plot, faceted by `major_category`. Which plot do you prefer and why?

## Which majors aren't that popular?

> 12. Which major category is the least popular in this sample? To answer this question we use a new function called `count`, which first groups the data and then counts the number of observations in each category (see below). Add to the pipeline appropriately to arrange the results so that the major with the lowest observations is on top.

```{r}
college_recent_grads %>%
  count(major_category)
```

## All STEM fields aren't the same

One of the sections of the [FiveThirtyEight story](https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/) is "All STEM fields aren't the same". Let's see if this is true.

First, let's create a new vector called `stem_categories` that lists the major categories that are considered STEM fields.

```{r}
stem_categories <- c("Biology & Life Science",
                     "Computers & Mathematics",
                     "Engineering",
                     "Physical Sciences")
```

> 13. Now, use this to create a new variable in our data frame indicating whether a major is STEM or not.

> 14. Which STEM majors have median salaries equal to or greater than the median for all majors's median earnings? Your output should only show the major name and median, 25th percentile, and 75th percentile earning for that major as and should be sorted such that the major with the highest median earning is on top.

## What types of majors do women tend to major in?

> 15. Create a scatterplot of median income vs. proportion of women in that major, colored by whether the major is in a STEM field or not. Describe the association between these three variables.

> 16. Which STEM majors have a higher proportion of women?

> 17. Which non-STEM majors have a higher proportion of women?
