---
title: 'Section 3: Data Wrangling (Part 1)'
author: 'MTH 365: Introduction to Data Science'
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

## Recommended Reading

- _Modern Data Science with R_ Ch. 4: Data Wrangling
- _Modern Data Science with R_ Ch. 5: Tidy Data and Iteration
- Wickham, Hadley. (2014). "Tidy Data". _Journal of Statistical Software_ 59(10). Available on BlueLine.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

## Data structure and semantics

- Most statistical datasets are tables made up of _rows_ and _columns_

A dataset is a collection of _values_: these can be _numbers_ (quantitative) or character _strings_ (qualitative)

- Every value belongs to a variable and an observation

__Variable__: 

__Observation__: 

## Tidy data

"Tidy" data is a standard way of mapping the meaning of a dataset to its structure. 

1. Each variable forms a column.
2. Each observation forms a row.
3. Each type of observational unit forms a table.

Any other arrangement of the data is called "messy".

person|treatment|result
---|---|---
John Smith|a|-
Jane Doe|a|16
Mary Johnson|a|3
John Smith|b|2
Jane Doe|b|11
Mary Johnson|b|1

Real datasets can, and often do, violate the three principles of tidy data in almost every way imaginable! 

## Example: Gapminder

Gapminder is an independent Swedish foundation with no political, religious or economic affiliations. __Gapminder is a fact tank, not a think tank.__ Gapminder fights devastating misconceptions about global development. Gapminder produces free teaching resources making the world understandable based on reliable statistics. Gapminder promotes a fact-based worldview everyone can understand.  Gapminder collaborates with universities, UN, public agencies and non-governmental organizations.

## Loading the Gapminder data set

```{r}
library(dslabs)
library(tidyverse)
data(gapminder)
glimpse(gapminder)
```

## Exploring the Gapminder data set

```{r, warning=FALSE}
ggplot(gapminder, aes(x=fertility, y=infant_mortality)) +
  geom_point()
```

```{r, warning=FALSE}
ggplot(gapminder, aes(x=fertility, y=infant_mortality)) +
  geom_point(aes(col=country))
```

```{r, warning=FALSE}
ggplot(gapminder, aes(x=fertility, y=infant_mortality)) +
  geom_point(aes(col=region))
```

```{r, warning=FALSE}
ggplot(gapminder, aes(x=fertility, y=infant_mortality)) +
  geom_point(aes(col=region)) + 
  facet_wrap(~year)
```

## Data wrangling with `dplyr`

There are five major `dplyr` functions for wrangling data:

- `select()`: take a subset of the columns (variables/features)
- `filter()`: take a subset of the rows (observations)
- `mutate()`: add or modify existing columns
- `arrange()`: sort the rows
- `summarize()`: aggregate the data across rows (grouping)

## `dplyr` rules for functions


## Piping v. layering

Pipe or layer?


```{r}
gapminder_short <- gapminder %>% select(gdp, region)
glimpse(gapminder_short)
```

## `filter()`

- `filter()`: 

```{r}
gapminder2000 <- gapminder %>% 
  filter(year==2000)
glimpse(gapminder2000)
```

```{r, warning=FALSE}
ggplot(gapminder2000, aes(x=fertility, y=infant_mortality)) +
  geom_point(aes(col=region))
```

```{r, eval=FALSE, warning=FALSE}
gapminder %>% filter(year==2000) %>%
  filter(region %in% c('Central Asia', 
                       'Eastern Asia', 
                       'Polynesia', 
                       'South-Eastern Asia', 
                       'Southern Asia', 
                       'Western Asia')) %>%
  ggplot(aes(x=fertility, y=infant_mortality)) +
  geom_point(aes(col=region))
```

```{r, warning=FALSE}
gapminder %>% filter(year==2000) %>%
  filter(region %in% c('Central Asia', 'Eastern Asia', 
                       'Polynesia', 'South-Eastern Asia', 
                       'Southern Asia', 'Western Asia')) %>%
  ggplot(aes(x=fertility, y=infant_mortality)) +
  geom_point(aes(col=region))
```

## (Smooth) Operators
Explain, in your own words, what these operations do.

- `%>%`: 
- `%in%`:
- `==`: 

## More Operators

Operator|Definition                   
---|---
`<`| less than                    
`<=`|	less than or equal to
`>`| greater than
`>=`|	greater than or equal to 
`==`|	exactly equal to 
`!=`|	not equal to 
`x & y`| `x` AND `y`                  
`x %in% y`| test if `x` is in `y`   


## Fertility and infant mortality in the US

Example: In your notes, modify your plot for another country. Describe how fertility and infant mortality are changing over time.

```{r, warning=FALSE}
gapminder %>% filter(country=='United States') %>%
  ggplot(aes(x=fertility, y=infant_mortality)) +
  geom_point(aes(col=year))
```

Relationship between fertility and infant mortality over time: 



## `mutate()`

- `mutate()` lets us create new variables based on manipulations of the old variables

Example: We'd like to calculate the gross domestic product per capita. Here are the variables in our data - write an expression to do this calculation.

```{r}
names(gapminder)
```

```{r}
gapminder <- gapminder %>% 
  mutate(GDP_pc = gdp/population)
```

Example: How do we find out which countries have the highest GDP per capita?

```{r, warning=FALSE, message=FALSE}
ggplot(gapminder, aes(x=GDP_pc))+geom_density()
```

Example: How do we find out which countries have the highest GDP per capita?

```{r}
high_GDP <- gapminder %>% filter(GDP_pc>50000)
```

```{r}
high_GDP
```

## `arrange()`

Right now, the data is sorted by country, then year. We could use the `arrange()` command to resort in terms of another variable.

```{r, eval=-2}
GDP_only <- gapminder %>% select(country, year, GDP_pc, continent)
GDP_only %>% arrange(desc(GDP_pc))
```

```{r, eval=FALSE}
GDP_only %>% arrange(continent, desc(GDP_pc))
```

## `summarize()`

If we want to compare summary statistics, we might use `summarize()`.

```{r}
GDP_only %>% summarize(avg_GDP_pc=mean(GDP_pc), 
                       sd_GDP_pc=sd(GDP_pc), 
                       min_GDP_pc=min(GDP_pc), 
                       max_GDP_pc=max(GDP_pc))
```

```{r}
GDP_only %>% filter(GDP_pc != "NA") %>%
  summarize(avg_GDP_pc=mean(GDP_pc), 
            sd_GDP_pc=sd(GDP_pc), 
            min_GDP_pc=min(GDP_pc), 
            max_GDP_pc=max(GDP_pc),
            N=n())
```

```{r}
GDP_only %>% filter(GDP_pc != "NA") %>%
  group_by(continent) %>%
  summarize(avg_GDP_pc=mean(GDP_pc), 
            sd_GDP_pc=sd(GDP_pc), 
            min_GDP_pc=min(GDP_pc), 
            max_GDP_pc=max(GDP_pc),
            N=n())
```

```{r}
GDP_only %>% filter(GDP_pc != "NA") %>%
  filter(year==2010) %>%
  group_by(continent) %>%
  summarize(avg_GDP_pc=mean(GDP_pc), 
            sd_GDP_pc=sd(GDP_pc), 
            min_GDP_pc=min(GDP_pc), 
            max_GDP_pc=max(GDP_pc),
            N=n())
```


Reflection Questions:

1.  With 1 being too slow and 5 being too fast, how was the speed of the video? 

2.  With 1 being too basic and 5 being too advanced, how was the level of the video? 

3.  Are there any topics you specifically want to address during the class meeting?

4.  Any additonal thoughts/comments on the notes:
